---
title: "Ordinal survey SEM with lavaan.survey"
author: "lavaan.survey contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
geometry: margin=1in
vignette: >
  %\VignetteIndexEntry{Ordinal survey SEM with lavaan.survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This vignette shows a complete ordinal CFA workflow with a complex sampling
design. The workflow has two stages:

1. Fit the ordinal model with `lavaan`, using `ordered = TRUE`.
2. Pass the fitted model and a `survey` design object to
   `lavaan.survey.ordinal()`.

The ordinal survey function estimates thresholds and polychoric correlations
with the sampling weights from the survey design, obtains their design-based
covariance matrix from replicate weights, and then refits the model with a
DWLS/WLSMV-style weight matrix.

## Setup

```{r packages}
library(MASS)
library(lavaan)
library(survey)
library(lavaan.survey)
```

## Simulate ordinal survey data

The example uses six ordinal indicators, two latent factors, clustering,
stratification, sampling weights, and a grouping variable. The same structure
also works with more indicators; a smaller example keeps the vignette quick to
run during package checks.

```{r simulate-data}
set.seed(42)

n <- 600
items <- paste0("y", 1:6)

Sigma <- matrix(0.20, nrow = 6, ncol = 6)
Sigma[1:3, 1:3] <- 0.60
Sigma[4:6, 4:6] <- 0.60
diag(Sigma) <- 1

continuous <- MASS::mvrnorm(n = n, mu = rep(0, 6), Sigma = Sigma)
dat <- as.data.frame(continuous)
names(dat) <- items

for (item in items) {
  dat[[item]] <- cut(
    dat[[item]],
    breaks = c(-Inf, -0.7, 0, 0.7, Inf),
    labels = 1:4,
    ordered_result = TRUE
  )
}

dat$cluster <- factor(rep(seq_len(60), each = n / 60))
dat$stratum <- factor(ifelse(as.integer(dat$cluster) <= 30, "A", "B"))
dat$weight <- round(runif(n, min = 0.5, max = 2.0), 2)
dat$gender <- factor(sample(c("girls", "boys"), n, replace = TRUE))
```

Declare the complex sampling design with the `survey` package.

```{r survey-design}
des <- survey::svydesign(
  id = ~cluster,
  strata = ~stratum,
  weights = ~weight,
  data = dat,
  nest = TRUE
)
```

## Single-group ordinal CFA

Start by fitting the ordinal CFA with `lavaan`. The important part is
`ordered = items`, which tells `lavaan` to estimate thresholds and polychoric
correlations for the ordinal indicators.

```{r model}
cfa_model <- '
  engagement =~ y1 + y2 + y3
  support    =~ y4 + y5 + y6
'
```

```{r single-group}
fit_naive <- lavaan::cfa(
  model = cfa_model,
  data = dat,
  ordered = items
)

fit_survey <- lavaan.survey.ordinal(
  lavaan.fit = fit_naive,
  survey.design = des,
  estimator = "WLSMV"
)
```

Selected robust fit measures can be extracted with the usual `lavaan`
functions.

```{r single-group-fit}
fitMeasures(
  fit_survey,
  c("chisq.scaled", "df.scaled", "pvalue.scaled", "cfi.robust", "rmsea.robust")
)
```

The parameter table also uses the usual `lavaan` interface.

```{r single-group-parameters}
subset(
  parameterEstimates(fit_survey, standardized = TRUE),
  op == "=~",
  select = c("lhs", "op", "rhs", "est", "se", "pvalue", "std.all")
)
```

## Multiple-group ordinal CFA

Multiple-group models follow the same pattern. First fit the grouping model in
`lavaan`, then apply the survey correction.

```{r multiple-group}
fit_configural <- lavaan::cfa(
  model = cfa_model,
  data = dat,
  ordered = items,
  group = "gender"
)

fit_configural_survey <- lavaan.survey.ordinal(
  lavaan.fit = fit_configural,
  survey.design = des,
  estimator = "WLSMV"
)

fitMeasures(
  fit_configural_survey,
  c("chisq.scaled", "df.scaled", "pvalue.scaled", "cfi.robust", "rmsea.robust")
)
```

## Loading and threshold invariance

For ordinal indicators, a common invariance step is to constrain factor
loadings and thresholds to equality across groups.

```{r scalar-invariance}
fit_scalar <- lavaan::cfa(
  model = cfa_model,
  data = dat,
  ordered = items,
  group = "gender",
  group.equal = c("loadings", "thresholds")
)

fit_scalar_survey <- lavaan.survey.ordinal(
  lavaan.fit = fit_scalar,
  survey.design = des,
  estimator = "WLSMV"
)

fitMeasures(
  fit_scalar_survey,
  c("chisq.scaled", "df.scaled", "pvalue.scaled", "cfi.robust", "rmsea.robust")
)
```

The equality constraints are preserved because `lavaan.survey.ordinal()`
refits the original `lavaan` parameter table.

```{r scalar-labels}
pe <- parameterEstimates(fit_scalar_survey)

subset(
  pe,
  op %in% c("=~", "|") & label != "",
  select = c("lhs", "op", "rhs", "group", "label", "est", "se")
)
```

## Practical notes

The current ordinal implementation is intended for standard `survey.design`
objects and replicate-weight designs supported by the `survey` package. If a
standard design is supplied, it is converted internally with
`survey::as.svrepdesign()`.

The ordinal variables used in the model should have stable categories in the
full sample and within each group. Very sparse categories can make thresholds
and polychoric correlations unstable, especially in multiple-group models.

At this stage, the ordinal extension is best validated for CFA/SEM models where
all observed model indicators are ordered categorical variables. Mixed
continuous and ordinal observed variables have an experimental single-group and
multiple-group path, including multiple imputation. Mixed models should be
treated as provisional until external validation is added.
