---
title: "Ordinal survey SEM with multiple imputation and groups"
author: "lavaan.survey contributors"
date: "`r Sys.Date()`"
output:
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
geometry: margin=1in
vignette: >
  %\VignetteIndexEntry{Ordinal survey SEM with multiple imputation and groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This vignette shows the combined workflow for ordered categorical indicators,
complex survey design information, multiple imputation, and multiple-group
models.

The analysis has four steps:

1. Create or obtain completed datasets from an imputation workflow.
2. Wrap the completed datasets in `mitools::imputationList()`.
3. Fit the ordinal multiple-group model once with `lavaan`.
4. Pass the fitted model and imputed survey design to
   `lavaan.survey.ordinal()`.

`lavaan.survey.ordinal()` then pools the ordinal sample statistics across
imputations: thresholds, polychoric correlations, and their design-based
covariance matrix.

## Setup

```{r packages}
library(lavaan)
library(survey)
library(mitools)
library(mice)
library(lavaan.survey)
```

## Simulate a small grouped ordinal survey dataset

The example uses four ordinal indicators, two groups, clusters, strata, and
sampling weights. The small size keeps the vignette quick to rebuild during
package checks; applied analyses should use larger samples and more imputations.

```{r simulate-data}
set.seed(20260513)

n_clusters <- 40
cluster_size <- 6
n <- n_clusters * cluster_size
items <- paste0("y", 1:4)

cluster <- rep(seq_len(n_clusters), each = cluster_size)
group_id <- rep(c(1L, 2L), each = n / 2)
group <- factor(group_id, levels = c(1L, 2L), labels = c("boys", "girls"))
stratum <- rep(rep(1:4, each = 5 * cluster_size), times = 2)

weight <- round(
  c(0.80, 0.95, 1.15, 1.35)[stratum] *
    ifelse(group == "boys", 1.05, 0.95) *
    runif(n, 0.85, 1.15),
  6
)

eta <- rnorm(n, mean = ifelse(group == "boys", -0.10, 0.20), sd = 1)
loadings <- c(0.82, 0.76, 0.72, 0.68)

ystar <- sapply(loadings, function(lambda) {
  lambda * eta + sqrt(1 - lambda^2) * rnorm(n)
})
colnames(ystar) <- items

to_ordered <- function(x) {
  ordered(
    cut(x, breaks = c(-Inf, -0.7, 0.1, 0.9, Inf), labels = 1:4),
    levels = 1:4
  )
}

dat <- as.data.frame(lapply(as.data.frame(ystar), to_ordered))
dat$group <- group
dat$group_id <- group_id
dat$weight <- weight
dat$cluster <- cluster
dat$stratum <- stratum
```

## Add missing ordinal responses

The missingness pattern is artificial but useful for demonstrating the workflow.
The imputation model below includes the grouping variable and survey-design
information. This is not a universal recipe; in real applications, the
imputation model needs substantive thought.

```{r missingness}
dat_missing <- dat

p_y2 <- plogis(-1.35 + 0.45 * scale(ystar[, "y1"])[, 1] +
                 0.25 * (dat$group == "girls"))
p_y4 <- plogis(-1.40 + 0.40 * scale(ystar[, "y3"])[, 1] -
                 0.20 * scale(dat$weight)[, 1])

dat_missing$y2[runif(n) < p_y2] <- NA
dat_missing$y4[runif(n) < p_y4] <- NA

colMeans(is.na(dat_missing[items]))
```

## Impute the ordered indicators

The example uses three imputations and three iterations to keep the vignette
fast. For applied analyses, use many more imputations, check convergence of the
imputation model, and document the imputation choices.

```{r imputation}
ini <- mice::mice(dat_missing, maxit = 0, printFlag = FALSE)
method <- ini$method
method[] <- ""
method[c("y2", "y4")] <- "polr"

predictor_matrix <- ini$predictorMatrix
predictor_matrix[,] <- 0
predictor_matrix[c("y2", "y4"), c(items, "group_id", "weight", "stratum")] <- 1
diag(predictor_matrix) <- 0

imp <- mice::mice(
  dat_missing,
  m = 3,
  maxit = 3,
  method = method,
  predictorMatrix = predictor_matrix,
  seed = 20260514,
  printFlag = FALSE
)

imputed_data <- lapply(seq_len(imp$m), function(i) {
  out <- mice::complete(imp, action = i)
  for (item in items) {
    out[[item]] <- ordered(out[[item]], levels = 1:4)
  }
  out$group <- factor(out$group, levels = c("boys", "girls"))
  out
})
```

## Build an imputed survey design

The `survey` package can store an imputation list as the design's data source.
This is the object that is passed to `lavaan.survey.ordinal()`.

```{r imputed-survey-design}
imp_list <- mitools::imputationList(imputed_data)

des_imp <- survey::svydesign(
  ids = ~cluster,
  strata = ~stratum,
  weights = ~weight,
  data = imp_list,
  nest = TRUE
)
```

## Fit a multiple-group ordinal CFA

First fit the model with `lavaan` on one completed dataset. The fitted model
contains the parameter table, group structure, ordered-variable information, and
any equality constraints that should be preserved.

```{r model}
cfa_model <- "
  f =~ y1 + y2 + y3 + y4
"
```

```{r lavaan-fit}
fit_naive <- lavaan::cfa(
  model = cfa_model,
  data = imputed_data[[1]],
  ordered = items,
  group = "group",
  group.equal = c("loadings", "thresholds"),
  estimator = "WLSMV"
)
```

## Apply the ordinal survey correction and MI pooling

Now pass the naive `lavaan` fit and the imputed survey design to
`lavaan.survey.ordinal()`.

```{r survey-fit}
fit_survey_imp <- lavaan.survey.ordinal(
  lavaan.fit = fit_naive,
  survey.design = des_imp,
  estimator = "WLSMV",
  rep.type = "bootstrap",
  replicates = 20
)
```

The returned object is a regular `lavaan` object, so the usual inspection
functions work.

```{r fit-measures}
fitMeasures(
  fit_survey_imp,
  c("chisq.scaled", "df.scaled", "pvalue.scaled",
    "cfi.robust", "tli.robust", "rmsea.robust", "srmr")
)
```

```{r loadings}
subset(
  parameterEstimates(fit_survey_imp, standardized = TRUE),
  op == "=~",
  select = c("group", "lhs", "op", "rhs", "est", "se", "pvalue", "std.all")
)
```

## Check that constraints are preserved

The equality labels from the original `lavaan` fit are preserved during the
survey-adjusted refit.

```{r constraint-labels}
pe <- parameterEstimates(fit_survey_imp)

subset(
  pe,
  op %in% c("=~", "|") & label != "",
  select = c("lhs", "op", "rhs", "group", "label", "est", "se")
)
```

## Practical notes

The vignette uses a deliberately small number of imputations and bootstrap
replicates to keep package checks fast. In applied research, increase both:
the number of imputations is a property of the missing-data problem, and the
number of replicate weights controls the stability of the design-based
covariance estimate.

For ordinal multiple-group models, sparse categories within groups are a common
source of instability. Before fitting the model, inspect category counts for
each ordered indicator by group and consider collapsing categories when needed.

Global fit indices under multiple imputation should be interpreted carefully.
`lavaan.survey.ordinal()` pools thresholds, polychoric correlations, and their
design-based covariance matrix before fitting one model; other software may
summarize fit across imputed-data analyses instead. Parameter estimates and
standard errors are usually the more direct cross-software comparison target.
